Layered Formal Verification of an Industrial TCP/IP Stack

### 2021-05-24 SecDev21/paper.tex
more rephrasing

@@

@@

Rewriting the part that process the incoming segments was out of the scope of
this work. Instead, we have opted for the use of symbolic execution on the
original CycloneTCP C code.
Symbolic execution is a mean to execute programs with symbolic values rather
than concrete values. Symbolic execution has already been successfully
used with other verification methods.
Indeed, in~\cite{vanoverberghe2008using} the authors present how to use symbolic
execution to improve deductive verification, and in~\cite{kassios2012comparing}
the authors explain that symbolic execution can be good at infering contracts.
For our work, KLEE has been used. It is built on top of LLVM as it takes LLVM
bytecode in input and works well with C code thanks to a very simple interface.

@@

The first step in order to extract contracts was to read the state machine to
know all the possible transitions that can be done by the reception of a flag,
and adapt it to
the subtleties of CycloneTCP. The next step is to write a driver to run KLEE and
know if the contracts that we have infered are correct or not.
A simplified example is shown in figure~\ref{code:kleedriver}. The first lines
create a symbolic incoming segment to process as well as a symbolic socket.
At line~\ref{code:kleedriver:tcpProcessSegment} the function that processes the
segment is called. It is followed at line~\ref{code:kleedriver:assert} by a KLEE
builtin that check if an assert is valid for all the symbolic paths
explored before.

@@

It appears that lot of programmers do not correctly use the socket API that is
provided by the library. Providing verified user functions for TCP is not enough
to guarantee the safety of software if the users do not correctly use them.
An important part of the work was to ensure that code that uses the socket API's
functions are correctly called. By ``correctly'' we mean: 1)~the functions
are called with respect to a defined order and 2)~the return code is checked
before continuing the processing.

@@

The TCP protocol implies a specific order such that the user can call the API's
functions without breaking the functional specification of the protocol. This
order also is being conveyed by the protocol's state machine given in
figure~\ref{Fig:statemachine}.
Ada's pre- and postconditions are a powerful tool to express such inter-function
dependencies, while SPARK technology can be used to guarantee that the
assertions always hold. Thus, pre- and postconditions were introduced to model
a partial order on the calls to the API's functions.

@@

Thanks to the pointer ownership policy adopted by SPARK and detailed in
section~\ref{sec:pointers}, we have found that a memory leak can happen when
the connection have been closed and the buffer are cleaning.

@@

Another very interesting and very subtle bug have been found thanks to SPARK is
reproduced in the snippet bellow.

@@

In this scenario, after the call to \spark{Tcp_Send}, when the error cases have
been filter, the state of the socket can either be \sestab or \sclosew. The call
at line~\ref{bugProg:tcpWaitForEvents} and now the state of the socket can be
still be \sestab or \sclosew, or if a \rst flag has been received when the mutex
was released, the state can be \sclosed too.
At line~\ref{bugProg:tcpWaitForEventsPost}, a \fin segment is sent, which does
not modify the state of the socket. Finally, at line~\ref{bugProg:tcpChangeState}
we try to jump to the state \sfwone and thus the transitions $\text{\sclosew}
\rightarrow \text{\sfwone}$ and $\text{\sclosed} \rightarrow \text{\sfwone}$ are
possible.

But these transition are not allowed by the norm, and thus not allowed by the
precondition of \spark{TCP_Change_State}. The problem is detected by SPARK and a fix
has been proposed.

@@

@@

The use of SPARK for the verification of the TCP protocol has given good
results. It has helped to find bugs in the existing implementation. But more
than finding bugs, the interest of this work lie in the face that it is a big
step toward a secure implementation. It has been proved by the tool GNATprove
that our implementation is free of run-time errors and that all the transitions
in the TCP state machine are done with respect to the norm.

@@

>>>>

% Yannick: the caption below used the term "user-task related function" which is used nowhere else

@@

% Yannick: the whole explanation is unclear. We should rephrase to explain that:
% - a suitable contract is added to function \spark{TCP_Wait_For_Events}
% - to make sure this contract is correct, we implement a ghost procedure
%   \spark{TCP_Wait_For_Events_Proof} as shown in Algorithm 1, that models
%   the possible chain of events that occur while waiting, with the same contract
%   as \spark{TCP_Wait_For_Events}. This procedure is proved automatically
%   without even a loop invariant thanks to loop unrolling in GNATprove.

@@

Rewriting the part that processes the incoming segments was out of the scope of
this work. Instead, we have opted for the use of symbolic execution on the
original CycloneTCP C code.
Symbolic execution is a mean to execute programs with symbolic values rather
than concrete values. Symbolic execution has already been successfully
used with other verification methods.
Indeed, in~\cite{vanoverberghe2008using} the authors present how to use symbolic
execution to improve deductive verification, and in~\cite{kassios2012comparing}
the authors explain that symbolic execution can be good at inferring contracts.
For our work, KLEE has been used. It is built on top of LLVM as it takes LLVM
bytecode in input and works well with C code thanks to a very simple interface.

@@

The first step in order to extract contracts was to read the state machine to
know all the possible transitions that can be done by the reception of a flag,
and adapt it to
the subtleties of CycloneTCP. The next step is to write a driver to run KLEE and
know if the contracts that we have inferred are correct or not.
A simplified example is shown in figure~\ref{code:kleedriver}. The first lines
create a symbolic incoming segment to process as well as a symbolic socket.
At line~\ref{code:kleedriver:tcpProcessSegment} the function that processes the
segment is called. It is followed at line~\ref{code:kleedriver:assert} by a KLEE
builtin that checks if an assert is valid for all the symbolic paths
explored before.

@@

A common problem faced by users of the CycloneTCP library is the need to
properly chain calls of the Socket API, in a way that respects the TCP
automaton. This entails verifying that 1)~the functions are called with respect
to a defined order and 2)~the return code is checked before continuing the
processing.

@@

The TCP protocol implies a specific order such that the user can call the API's
functions without breaking the functional specification of the protocol. This
order also is being conveyed by the protocol's state machine given in
figure~\ref{Fig:statemachine}.
Ada's preconditions and postconditions are a powerful tool to express such inter-function
dependencies, while SPARK technology can be used to guarantee that the
assertions always hold. Thus, preconditions and postconditions were introduced to model
a partial order on the calls to the API's functions.

@@

Thanks to the pointer ownership policy adopted by SPARK and detailed in
section~\ref{sec:pointers}, we have found that a memory leak can happen when
the connection has been closed and the buffer are cleaning.

@@

A subtle bug found thanks to SPARK is
reproduced in the snippet below.

@@

In this scenario, after the call to \spark{Tcp_Send}, when the error cases have
been filtered, the state of the socket can either be \sestab or \sclosew. After the call
at line~\ref{bugProg:tcpWaitForEvents}, the state of the socket can
still be \sestab or \sclosew, or it can be \sclosed if a \rst flag has been received when the mutex
was released.
At line~\ref{bugProg:tcpWaitForEventsPost}, a \fin segment is sent, which does
not modify the state of the socket. Finally, at line~\ref{bugProg:tcpChangeState}
we try to jump to the state \sfwone and thus the transitions $\text{\sclosew}
\rightarrow \text{\sfwone}$ and $\text{\sclosed} \rightarrow \text{\sfwone}$ are
possible.

But these transitions are not allowed by RFC 793, and thus not allowed by the
precondition of \spark{TCP_Change_State}. The problem is detected by SPARK and a fix
has been applied to the SPARK and C implementations of CycloneTCP.

% The above is uncler, the transitions from CLOSED and CLOSED-WAIT are allowed, it's only the transition from ESTABLISHED which is not allowed.

@@

% above: need time to prove automatically with GNATprove with the spec of your laptop (num of cores, amount of memory)

@@

The use of SPARK for the verification of the TCP protocol has given good
results. It has helped to find bugs in the existing implementation. But more
than finding bugs, the interest of this work lie in the face that it is a
step toward a secure implementation. It has been proved by the tool GNATprove
that our implementation is free of run-time errors and that all the transitions
in the TCP state machine are done with respect to the norm.

@@

% examples of such properties?

% ambition for industrial use?
